---
title: "Biostatistics Toolkit 4: Plotting Data"
output:
  html_document:
    df_print: paged
---

## Introduction 

In this last notebook of today's workshop, you'll get more practice reading in data and make some cool plots in `ggplot`. As before, anything with an `<XXX>` requires an answer from you. 

As always, let's make sure you've loaded the tidyverse. 

```{r}
library(tidyverse);
```


In your directory is a text (.txt) file named 'rats_info.txt'. It quotes from Section 2.1 of Verbeke and Molenberghs (2009):

*The primary aim was to investigate the effect of inhibition of the production of testosterone in male Wistar Rats on their craniofacial growth.*

*A total of 50 male Wistar rats have been randomized to either a control group or one of the two treatment groups where treatment consisted of a low or high dose of the drug Decapeptyl, which is an inhibitor for testosterone production in rats. The treatment started at the age of 45 days, and measurements were taken every 10 days, with the first observation taken at the age of 50 days. The responses of interest are distances (in pixels) between well-defined points on X-ray pictures of the skull of each rat, taken after the rat has been anesthetized. Of primary interest is the estimation of changes over time and testing whether these changes are treatment dependent.*

*For the purpose of this book, we will consider one of the measurements which can be used to characterize the height of the skull.*

Note that there are just 50 individual rats but that each rat has been observed at multiple timepoints. Also in your directory is a comma-separated value (.csv) file called 'rats.csv'.  Let's read in the data now. 

```{r}
rat_data <- read_csv("rats.csv");
```

After running this chunk, you should have received an informational message from `R` that begins `Parsed with column specification...` In our previous notebook (`03-reading_data.Rmd`), I had you provide explicit instructions as to what type of data (real numbers, integers, characters, etc) were in each column of the birthweight dataset. We didn't provide any explicit instructions here, and `R` seems to have made some decisions for you. Let's call up the data and see how `R` understands the data:

```{r}
rat_data;
```

`R` has interpreted the observation number, the rat ID, and age (in days) of the rat at each observation as integers; the treatment label is a character; and the response is a double (a real number). These decisions are acceptable with one exception. Specifically, I would say that the rat ID is not a true integer, e.g. rat \# 5 is not greater than rat \# 3, but should rather be considered a *character*. Re-read in the data using the following chunk, replacing the `<XXX>` with the appropriate value. 

```{r}
rat_data <- read_csv("rats.csv",
                     col_types = cols(rat = col_<XXX>()));
```

## Spaghettti plots

These data are slightly more complicated than the birthweight data, because of the longitudinal aspect of the data. Let's try something we know first. Run the following two chunks (the first might take `R` a little while to create):

```{r}
ggplot(rat_data) +
  geom_point(aes(x = age, y = response, color = rat));
```

```{r}
ggplot(rat_data) +
  geom_point(aes(x = age, y = response, color = rat)) + 
  guides(color = FALSE);
```

Based on this, what does the `+ guides(color = FALSE)` do? Which plot is more useful in your opinion?

<XXX>

What crucial aspect of the data are you not able to see in either of the above plots? 

<XXX>

There's a geom for connecting the dots, and you might even surmise its name: `geom_line`. Type the following code in your console to pull up its help page: `?geom_line`. As you type, notice how `R`studio pulls up suggestions for what you might be typing, similar to what Google does with its search engine. If you ever see what you're looking for in the list of suggestions, navigate to it using the up- or down-arrows and press 'enter' to fill it in. 

Reading the help page, we see that there are actually several options to choose from. The first two sentences of the help page for `geom_line` are as follows: *geom_path() connects the observations in the order in which they appear in the data. geom_line() connects them in order of the variable on the x axis.* Based on this, which one of these are we interested in? 

<XXX>

```{r}
ggplot(rat_data) +
  geom_<XXX>(aes(x = age, y = response, colour = rat)) + 
  guides(color = FALSE);
```

If you've gotten this far, congratulations: you've created your first "spaghetti" plot! Each strand of spaghetti represents a different rat. Buon appetito!


## Getting fancy 

Color might not be the best aesthetic to distinguish between rats. Sometimes you just want to tell `ggplot` that there are different groups, i.e. rats, without making them visually distinctive. There's an aesthetic for that! Run the following (after replacing <XXX>)

```{r}
ggplot(rat_data) +
  geom_<XXX>(aes(x = age, y = response, group = rat));
```

And now color has been freed up for use with another variable, like treatment:

```{r}
ggplot(rat_data) +
  geom_<XXX>(aes(x = age, y = response, group = rat, color = treat));
```

Note: now we *want* the legend for color. 

## Even fancier

Let's apply some `tidyverse` expertise to create some nice visual summaries. Run the following chunk:

```{r}
rat_data <- 
  rat_data %>%
  group_by(treat, age) %>%
  mutate(mean_response = mean(response)) %>%
  ungroup();
```

What does the above code chunk do?

<XXX>

Run the following chunk

```{r}
ggplot(rat_data) +
  geom_<XXX>(aes(x = age, y = mean_response, group = rat, color = treat), size = 1.5);
```

Look carefully at the above chunk: `size` is *outside* of the call to `aes()`, and it has an explicit value of 1.5 and not a variable name. An aesthetic links a visual characteristic to the value of a variable, so technically `size` is not an aesthetic in this context. 

Now, make your own plot, trying to include as many of the following characteristics as possible.  

(i) each individual mouse spaghetti is plotted, colored by treatment group
(ii) the average spaghetti over all rats in each treatment group is plotted, and the size of the is set to 1.5
(iii) in order to minimize the visual impact of any given rat's spaghetti, make it slightly transparent by putting `alpha = 0.5` at the right place in the right geom. 

Hint: you've done most of the work already, you just need to "add" the right geoms together and tweak a few things.

```{r}
#<XXX>
```


## Putting it all together

Don't forget to click the *Knit* button or press *Cmd+Shift+K* (*Ctrl+Shift+K*) to create an HTML file of your entire notebook. Congratulations! You're done with four notebooks!

## References

Verbeke, Geert, and Geert Molenberghs. *Linear mixed models for longitudinal data*. Springer Science & Business Media, 2009.
